---
title: "Analysis of Gapminder Dataset using R"
author: "Catriona McIntosh"
date: 2026-09-23
---
```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(gapminder)
```

This post uses the gapminder dataset to look at life expectancy across 142 different countries from 1952 to 2007

Data: Based on free material from [GAPMINDER.ORG](https://www.gapminder.org/data/),
CC-BY licence, via the [gapminder R package](https://cran.r-project.org/package=gapminder)
(Jennifer Bryan).

## Life Expectancy by continent

I took the averaged life expectancy across the countries in each continent for every survey year (1952-2007) to compare how life expectancy has changed in each continent.

```{r}
continent_trend <- gapminder |>
  group_by(continent, year) |>
  summarise(mean_life = mean(lifeExp), .groups = "drop")
```

```{r}
#| label: fig-continent-trend
#| fig-cap: "Mean life expectancy by continent, 1952–2007."
ggplot(continent_trend, aes(year, mean_life, colour = continent)) +
  geom_line(linewidth = 1) +
  geom_point() +
  labs(x = NULL, y = "Mean life expectancy (years)", colour = NULL) 
```

As @fig-continent-trend shows that life expectancy has consistently been highest in Oceania, and lowest in Africa. Each continent has seen an increase in life expectancy between 1952 and 2007, with each continent's life expectancy increasing monotonically except for Africa. There was a dip in life expectancy in Africa between the late 1990s and early 2000s before increasing again.

## Which country gained the most?

Next I calculated the change in life expectancy for each country between
1952 and 2007.

```{r}
life_change <- gapminder |>
  filter(year %in% c(1952, 2007)) |>
  select(country, continent, year, lifeExp) |>
  pivot_wider(names_from = year, values_from = lifeExp, names_prefix = "y") |>
  mutate(change = y2007 - y1952) |>
  arrange(desc(change))
 
n_declined <- sum(life_change$change < 0)
top <- slice_head(life_change, n = 1)
bottom <- slice_tail(life_change, n = 1)
```

```{r}
#| label: life-change-extremes
#| fig-cap: "Top/Bottom 5 changes in life expectancy"
bind_rows(slice_head(life_change, n = 5), slice_tail(life_change, n = 5))
```

`r top$country` gained the most, `r round(top$change, 1)` years. At the other end, `r n_declined` countries had a *lower*
life expectancy in 2007 than in 1952. The largest drop was in
`r bottom$country` (`r round(bottom$change, 1)` years).
